---
title: "COVID-19 Mortality & Long-Term Air Pollution Exposure"
description: |
  Detailled Script.
author:
  - name: Léo Zabrocki 
    url: https://www.parisschoolofeconomics.eu/
    affiliation: Paris School of Economics
    affiliation_url: https://www.parisschoolofeconomics.eu/
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    keep_md: true
    toc: true
    toc_depth: 2
---

<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
# code chunk option
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  layout="l-body-outset",
  dev = "CairoPNG",
  dpi = 300
)
```

In this document, we explore the relationship between long-term air pollution exposure and COVID-19 mortality. Should you have any questions or find coding errors, please do not hesitate to reach me at **leo.zabrocki@psemail.eu**.

# Required Packages and Loading Data

To reproduce exactly the `script_mortality_long_term_air_pollution.html` document, you first need to have installed:

* the [R](https://www.r-project.org/) programming language on your computer 
* [RStudio](https://rstudio.com/), an integrated development environment for R, which will allow you to knit the `script_mortality_long_term_air_pollution.Rmd` file and interact with the R code chunks
* the [R Markdown](https://rmarkdown.rstudio.com/) package
* and the [Distill](https://rstudio.github.io/distill/) package which provides the template of this document. 

Once everything is set up, we need to load the following packages:

```{r}
# load required packages
library(rmarkdown) # for creating the R Markdown document
library(knitr) # for creating the R Markdown document
library(here) # for files paths organization
library(tidyverse) # for data manipulation and visualization
library(modelr)     # provides easy pipeline modeling functions
library(broom)      # helps to tidy up model outputs
library(lubridate) # for manipulating date variables
library(Cairo) # for printing customed police of graphs
```

We also load our custom `ggplot2` theme for graphs:

```{r}
# load ggplot custom theme
source(
  here::here(
    "2.scripts",
    "4.custom_ggplot2_theme",
    "script_custom_ggplot_theme.R"
  )
)
```

The theme is based on the fantastic [hrbrthemes](https://hrbrmstr.github.io/hrbrthemes/index.html) package. If you do not want to use this theme or are unable to install it because of fonts issues, you can use the `theme_mimimal()` included in the `ggplot2` package.

We load the data we use in our analysis:

```{r}
# load data
data <-
  readRDS(here::here("1.data", "3.data_for_analysis", "data_analysis.rds"))
```

# Data Wrangling

We select the relevant variables:

```{r}
data <- data %>%
  dplyr::select(
    city,
    patient_id,
    covid:cancer,
    mean_2017_2019_no2:mean_2017_2019_so2,
    mean_2017_2019_average_temperature,
    mean_2017_2019_relative_humidity
  )
```

We keep patients that were COVID-positive:

```{r}
data <- data %>%
  filter(covid == 1) %>%
  dplyr::select(-covid)
```

# Logistic Regression Analysis

Models:

```{r}
data_regression_analysis <- data %>%
  pivot_longer(
    cols = c(mean_2017_2019_no2:mean_2017_2019_pm2p5),
    names_to = "pollutant",
    values_to = "concentration"
  ) %>%
  group_by(pollutant) %>%
  nest()

data_regression_analysis <- data_regression_analysis %>%
  mutate(
    # model with only concentration
    `Model Only With Concentration` = map(data, ~ glm(dead ~ concentration, family = "binomial",
                                                         data = .)),
    # model with concentration, city and all patient characteristics
    `Model With Concentration, All Patient Characteristics\nand City Fixed-Effects` = map(
      data,
      ~ glm(
        dead ~ concentration +
          sex +
          age +
          bmi +
          diabetes +
          hbp +
          copd +
          cardiac_disease_infarction +
          renal_failure +
          cancer +
          city,
        family = "binomial",
        data = .
      )
    ),
    # model with concentration, all patient characteristics but without city
    `Model With Concentration and All Patient Characteristics` = map(
      data,
      ~ glm(
        dead ~ concentration +
          sex +
          age +
          bmi +
          diabetes +
          hbp +
          copd +
          cardiac_disease_infarction +
          renal_failure +
          cancer,
        family = "binomial",
        data = .
      )
    )
  )

# transform in long according to models
data_regression_analysis <- data_regression_analysis %>%
  pivot_longer(
    cols = c(
      `Model Only With Concentration`:`Model With Concentration and All Patient Characteristics`
    ),
    names_to = "model",
    values_to = "coefficients"
  ) %>%
  dplyr::select(-data)

# tidy regression ouputs
data_regression_analysis <- data_regression_analysis %>%
  mutate(models_dfs = map(coefficients, ~ broom::tidy(., conf.int = TRUE)))

# unnest results and select coefficient for total gross tonnage
data_regression_analysis <- data_regression_analysis %>%
  unnest(models_dfs) %>%
  dplyr::filter(term == "concentration")

# exponentiate
data_regression_analysis <- data_regression_analysis %>%
  mutate_at(vars(estimate, conf.low, conf.high), ~ exp(.))
```

Graph for the three models:

```{r, fig.width=30, fig.height=15}
# make the graph
graph_mortality_air_pollution_all_models <- data_regression_analysis %>%
  mutate(
    pollutant = case_when(
      pollutant == "mean_2017_2019_no2" ~ "NO2 (µg/m3)",
      pollutant == "mean_2017_2019_o3" ~ "O3 (µg/m3)",
      pollutant == "mean_2017_2019_pm10" ~ "PM10 (µg/m3)",
      pollutant == "mean_2017_2019_pm2p5" ~ "PM2.5 (µg/m3)",
      pollutant == "mean_2017_2019_so2" ~ "SO2 (µg/m3)"
    )
  ) %>%
  mutate(
    model = fct_relevel(
      model,
      "Model Only With Concentration",
      "Model With Concentration and All Patient Characteristics",
      "Model With Concentration, All Patient Characteristics\nand City Fixed-Effects")
  ) %>%
  ggplot(., aes(
    x = estimate,
    y = fct_rev(pollutant),
    xmin = conf.low,
    xmax = conf.high
  )) +
  geom_vline(xintercept = 1) +
  geom_pointrange(size = 1.2, colour = "deepskyblue3") +
  facet_wrap( ~ model, nrow = 1) +
  ylab("") + xlab("Odds of Dying from COVID-19") +
  custom_theme +
  theme(plot.margin = margin(1, 1, 1, -1, "cm"))


# print the graph
graph_mortality_air_pollution_all_models

# save graph
ggsave(graph_mortality_air_pollution_all_models, filename = here::here("3.outputs", "1.figures", "3.study_mortality_long_term_air_pollution", "graph_mortality_air_pollution_all_models.pdf"), 
       width = 60, height = 15, units = "cm", device = cairo_pdf)
```

Graph for the full model:

```{r, fig.width=20, fig.height=8}
# make the graph
graph_mortality_air_pollution_full_model <- data_regression_analysis %>%
  mutate(
    pollutant = case_when(
      pollutant == "mean_2017_2019_no2" ~ "NO2 (µg/m3)",
      pollutant == "mean_2017_2019_o3" ~ "O3 (µg/m3)",
      pollutant == "mean_2017_2019_pm10" ~ "PM10 (µg/m3)",
      pollutant == "mean_2017_2019_pm2p5" ~ "PM2.5 (µg/m3)",
      pollutant == "mean_2017_2019_so2" ~ "SO2 (µg/m3)"
    )
  ) %>%
  filter(model == "Model With Concentration, All Patient Characteristics\nand City Fixed-Effects") %>%
  ggplot(., aes(
    x = estimate,
    y = fct_rev(pollutant),
    xmin = conf.low,
    xmax = conf.high
  )) +
  geom_vline(xintercept = 1) +
  geom_pointrange(size = 1.2, colour = "deepskyblue3") +
  ylab("") + xlab("Odds of Dying from COVID-19") +
  custom_theme +
  theme(plot.margin = margin(1, 1, 1, -1, "cm"))


# print the graph
graph_mortality_air_pollution_full_model

# save graph
ggsave(graph_mortality_air_pollution_full_model, filename = here::here("3.outputs", "1.figures", "3.study_mortality_long_term_air_pollution", "graph_mortality_air_pollution_full_model.pdf"), 
       width = 30, height = 15, units = "cm", device = cairo_pdf)
```


Table for the full model:

```{r}
# print the table
data_regression_analysis %>%
  mutate(
    pollutant = case_when(
      pollutant == "mean_2017_2019_no2" ~ "NO2 (µg/m3)",
      pollutant == "mean_2017_2019_o3" ~ "O3 (µg/m3)",
      pollutant == "mean_2017_2019_pm10" ~ "PM10 (µg/m3)",
      pollutant == "mean_2017_2019_pm2p5" ~ "PM2.5 (µg/m3)",
      pollutant == "mean_2017_2019_so2" ~ "SO2 (µg/m3)"
    )
  ) %>%
  filter(model == "Model With Concentration, All Patient Characteristics\nand City Fixed-Effects") %>%
  dplyr::select(pollutant, estimate, conf.low, conf.high) %>%
  mutate_at(vars(estimate:conf.high), ~ round(., 2)) %>%
  rename("Pollutant" = pollutant, "Point Estimate" = estimate, "Lower Bound 95% CI" = conf.low, "Upper Bound 95% CI" = conf.high) %>%
  kable(align = c("l", "l", rep("c", 3)))
```