---
title: "Data Wrangling"
description: |
  Merging Patients' Data with Air Pollution Exposures.
author:
  - name: Léo Zabrocki 
    url: https://www.parisschoolofeconomics.eu/
    affiliation: Paris School of Economics
    affiliation_url: https://www.parisschoolofeconomics.eu/
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---

<style>
body {
text-align: justify}
</style>


This script merges patients' COVID data for Bordeaux, Bruxelles, La Roche-sur-Yon, Nantes, Paris, Reims, Strasbourg, and Tours with their air pollution exposures computed by the Copernicus Atmosphere Monitoring Service (CAMS). Should you have any questions or find coding errors, please do not hesitate to reach me at **leo.zabrocki@psemail.eu**.

# Required Packages

To reproduce exactly the `script_data_wrangling.html` document, you first need to have installed:

* the [R](https://www.r-project.org/) programming language on your computer 
* [RStudio](https://rstudio.com/), an integrated development environment for R, which will allow you to knit the `script_data_wrangling.Rmd` file and interact with the R code chunks
* the [R Markdown](https://rmarkdown.rstudio.com/) package
* and the [Distill](https://rstudio.github.io/distill/) package which provides the template of this document. 

Once everything is set up, we need to load the following packages:

```{r, echo=TRUE, message = FALSE, warning = FALSE}
# load required packages
library(rmarkdown) # for creating the R Markdown document
library(knitr) # for creating the R Markdown document
library(here) # for files paths organization
library(tidyverse) # for data manipulation and visualization
library(data.table) # for loading heavy data
library(lubridate) # for manipulating date variables
```

# Loading and Cleaning COVID Data

### Data Loading

We first load the data located in the `1.data\1.covid_data` folder:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# we get the path of each file
covid_data <-
  tibble(file = list.files(
    path = here::here("1.data", "1.covid_data", "2.clean_data"),
    pattern = ".csv",
    full.names = T
  )) %>%
  # we open each file using a nested data frame structure
  mutate(data = map(file, ~ fread(.))) %>%
  # convert all variables to characters
  mutate(data = map(data, ~ mutate_all(.x, ~ as.character(.)))) %>%
  # unnest each city data
  unnest(data) %>%
  # from the files paths, get the city name
  mutate(city = basename(file) %>%
           word(., 1, sep = "\\_")) %>%
  # drop the file variable
  select(-file)
```

### Cleaning the Data

We clean the data:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
covid_data <- covid_data %>%
  # set the name of patients to lower cases and correct labels for bordeaux and bruxelles
  mutate(patient_id = tolower(patient_id)) %>%
  mutate(
    patient_id = ifelse(
      city == "bordeaux",
      str_replace_all(patient_id, "b", "bo"),
      patient_id
    ) %>%
      ifelse(city == "bruxelles", str_replace_all(patient_id, "b", "br"), .)
  ) %>%
  # label the sex variable
  mutate(sex = case_when(sex == "1" ~ "male",
                         sex %in% c(0, 2) ~ "female")) %>%
  # recode covid variable
  mutate(
    covid = case_when(
      covid == "Positief voor SARS-CoV-2 (COVID-19)" ~ 1,
      covid == "Negatief voor SARS-CoV-2 (COVID-19)" ~ 0,
      covid == "0" ~ 0,
      covid == "1" ~ 1
    )
  ) %>%
  # recode bmi variable
  mutate(
    bmi = ifelse(bmi == "ND", NA, bmi) %>%
      ifelse(bmi == "22,4 en nov19", 22.4, .) %>%
      ifelse(bmi == "20,4 en oct19", 20.4, .) %>%
      ifelse(bmi == "14,5 en janv20", 14.5, .) %>%
      ifelse(bmi == "19,2 en 11/2018", 19.2, .) %>%
      ifelse(bmi == "18,4 le 09/03/20", 18.4, .) %>%
      ifelse(bmi == "23,3 le 18/03/20", 23.3, .) %>%
      ifelse(bmi == "29,5 en 03/2020", 29.5, .) %>%
      ifelse(bmi == "?", NA, .)
  ) %>%
  mutate(bmi = as.numeric(sub(",", ".", bmi, fixed = TRUE))) %>%
  # compute age for strasbourg patients' id s1-s95
  mutate(age = ifelse(patient_id %in% paste("s", seq(1:95), sep = ""), 2020 - year(mdy(date_birth)), age)) %>%
  # drop date_birth variable
  select(-date_birth) %>%
  # set the pcr date as a date variable
  mutate(date_pcr_test = lubridate::mdy(date_pcr_test)) %>%
  # we have to drop bruxelles patient id 197 because it is duplicated
  # and we do not know for which patient copernicus computed the air pollution exposure
  filter(patient_id != "br197")
```

We check the coding of the `covid` variable:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# values
unique(covid_data$covid)
# type
typeof(covid_data$covid)
```

We check the coding of the `dead` variable:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# values
unique(covid_data$dead)
# type
typeof(covid_data$dead)
```

We convert the variable to numeric:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
covid_data$dead <- as.numeric(covid_data$dead)
```

We check the coding of the `sex` variable:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# values
unique(covid_data$sex)
# type
typeof(covid_data$sex)
```

We check the coding of the `age` variable:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# values
unique(covid_data$age)
# type
typeof(covid_data$age)
```

We convert the variable to numeric:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
covid_data$age <- as.numeric(covid_data$age)
```

We check the coding of the `hbp` variable:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# values
unique(covid_data$hbp)
# type
typeof(covid_data$hbp)
```

We convert the variable to numeric:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
covid_data$hbp <- as.numeric(covid_data$hbp)
```

We check the coding of the `diabetes` variable:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# values
unique(covid_data$diabetes)
# type
typeof(covid_data$diabetes)
```

We convert the variable to numeric:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
covid_data$diabetes <- as.numeric(covid_data$diabetes)
```

We check the coding of the `active_smoking` variable:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# values
unique(covid_data$active_smoking)
# type
typeof(covid_data$active_smoking)
```

We convert the variable to numeric:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
covid_data$active_smoking <- as.numeric(covid_data$active_smoking)
```

We check the coding of the `active_past_smoking` variable:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# values
unique(covid_data$active_past_smoking)
# type
typeof(covid_data$active_past_smoking)
```

We convert the variable to numeric:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
covid_data <- covid_data %>%
  mutate(
    active_past_smoking = ifelse(
      active_past_smoking %in% c("ancien sevré", "1 sevré", "sevré"),
      1,
      active_past_smoking
    ) %>%
      as.numeric(.)
  )
```

We check the coding of the `cardiac_disease_infarction` variable:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# values
unique(covid_data$cardiac_disease_infarction)
# type
typeof(covid_data$cardiac_disease_infarction)
```

We convert the variable to numeric:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
covid_data$cardiac_disease_infarction <- as.numeric(covid_data$cardiac_disease_infarction)
```

We check the coding of the `copd` variable:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# values
unique(covid_data$copd)
# type
typeof(covid_data$copd)
```

We convert the variable to numeric:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
covid_data$copd <- as.numeric(covid_data$copd)
```

We check the coding of the `renal_failure` variable:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# values
unique(covid_data$renal_failure)
# type
typeof(covid_data$renal_failure)
```

We convert the variable to numeric:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
covid_data$renal_failure <- as.numeric(covid_data$renal_failure)
```

We check the coding of the `cancer` variable:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# values
unique(covid_data$cancer)
# type
typeof(covid_data$cancer)
```

We convert the variable to numeric:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
covid_data <- covid_data %>%
  mutate(cancer = ifelse(cancer %in% c("2", "3", "5"), 1, cancer) %>%
           as.numeric(.))
```

We check the coding of the `bmi` variable:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# type
typeof(covid_data$bmi)
```

### English Translation of Variables

We provide below the translation of French names into English names that we use:

* "nom" ou "code patient": `patient_id`.
* "sexe": `sex`
* "age": `age`
* "décès": `dead`
* "HTA": `hbp`
* "diabète": `diabetes`
* "cardiopathie, cardio ischémique, atcd infarctus": `cardiac_disease_infarction`
* "BPCO": `copd`
* "insuffisance rénale chronique": `renal_failure`
* "Date RT-PCR": `date_pcr_test`
* "cancers": `cancer`
* "imc": `bmi`
* "tabagisme actif": `active_smoking`
* "tabagisme actif ou ancien": `active_past_smoking`

There are additional variables in Bruxelles data:

* "tabagisme ancien" : `past_smoking`
* "alcool": `alcohol`
* "hypercholestérolemie": `hypercholesterolemia`
* "CVA/TIA": `cva_tia`
* "mal vasculaire périphérique": `peripheral_vascular_disease`
* "cockcroft": `cockcroft`
* "nombre de lobes atteints": `n_affected_lobes`
* "Score verre dépoli /25": `score_verre_depoli_25`
* "intubation": `intubation`
* "soins intensifs": `intensive_care`
* "durée de séjour": `length_stay`
* "réhospitalisation": `rehospitalization`     

It is important to note that smoking is not coded in the same maner for Strasbourg data since the variable for smoking (`active_past_smoking`) does not distinguish current from past smoking.

### Selecting Relevant Variables

We select the set of variables that are common to all cities:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# select and reorder relevant variables
covid_data <- covid_data %>%
  select(
    # patient city and id
    city,
    patient_id,
    # covid outcomes
    date_pcr_test,
    covid,
    dead,
    # patient sex, age, and bmi
    sex,
    age,
    bmi,
    # patient smoking
    active_smoking,
    active_past_smoking,
    # patient diseases
    hbp,
    diabetes,
    cardiac_disease_infarction,
    copd,
    renal_failure,
    cancer
  )
```

The dataset contains `r nrow(covid_data)` observations and `r ncol(covid_data)` variables.

# Loading and Cleaning Air Pollution Data

Copernicus Atmosphere Monitoring Service (CAMS) provided us with two air pollution exposures files (2017-2019 and 2020) for each patient.

### Gathering Air Pollution Data in a Single File

We first gather all individuals files into a unique file and do some data cleaning:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# we get the path of each file
air_pollution_data <-
  tibble(file = list.files(
    path = here::here("1.data", "2.air_pollution_data"),
    pattern = ".csv",
    full.names = T
  )) %>%
  # we open each file using a nested data frame structure
  mutate(data = map(file, ~ fread(.))) %>%
  # we get the patient id from the files' names
  mutate(patient_id = basename(file) %>%
           word(., 2, sep = "\\_") %>%
           tolower(.)) %>%
  # we drop the "a" letter from patient id as it indicates the exposure of a patient's second adress
  # we will take the mean of exposures for the two adresses
  mutate(patient_id = str_remove(patient_id, "a")) %>%
  # we get the names of cities from the patient id
  mutate(
    city = case_when(
      str_detect(str_sub(
        patient_id, start = 1, end = 2
      ), "bo") ~ "bordeaux",
      str_detect(str_sub(
        patient_id, start = 1, end = 2
      ), "br") ~ "bruxelles",
      str_detect(str_sub(
        patient_id, start = 1, end = 1
      ), "n") ~ "nantes",
      str_detect(str_sub(
        patient_id, start = 1, end = 1
      ), "p") ~ "paris",
      str_detect(str_sub(
        patient_id, start = 1, end = 1
      ), "r") ~ "reims",
      str_detect(str_sub(
        patient_id, start = 1, end = 1
      ), "s") ~ "strasbourg",
      str_detect(str_sub(
        patient_id, start = 1, end = 1
      ), "t") ~ "tours",
      str_detect(str_sub(
        patient_id, start = 1, end = 1
      ), "v") ~ "roche"
    )
  ) %>%
  # we drop files' names
  select(-file) %>%
  # we unnest each air pollution data
  unnest(data) %>%
  # we rename basetime as date
  rename(date = basetime) %>%
  # we set the date variable as date
  mutate(date = lubridate::ymd(date)) %>%
  # we create the year variable
  mutate(year = lubridate::year(date)) %>%
  # we rename some variables
  rename(
    wind_speed = `10ff`,
    boundary_layer_height = blh,
    relative_humidity = rh,
    average_temperature = t2m
  ) %>%
  # we reorder the variables
  select(
    # patient id and city
    patient_id,
    city,
    # data and year
    date,
    year,
    # pollutant concentrations
    no2,
    o3,
    pm10,
    pm2p5,
    so2,
    # weather variables
    average_temperature,
    relative_humidity,
    wind_speed,
    boundary_layer_height,
    uvi,
    # pollen
    birch,
    grass,
    olive
  )
```

We then check if the Copernicus team computed the exposure of all patients:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
covid_data %>%
  filter(!(patient_id %in% air_pollution_data$patient_id)) %>%
  pull(patient_id)
```

We can see that three patients with positive PCR tests and several patients with negative PCR tests in Strasbourg do not have exposure data. We can also check if Copernicus computed the exposure for patients for which we do not have health data:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
air_pollution_data %>%
  # we drop patients ids which are not in the covid data
  filter(!(patient_id %in% covid_data$patient_id)) %>%
  pull(patient_id) %>%
  unique()
```
Copernicus did compute exposure of patients which were removed from the COVID datasets.

### Computing Average Exposures for 2017-2019

We then compute the average exposure for each patient over the 2017-2019 period:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
air_pollution_average_exposure_2017_2019 <- air_pollution_data %>%
  # group by patient
  group_by(patient_id) %>%
  # compute the long-term exposure before the pcr test
  filter(year < 2020) %>%
  # compute the mean of air pollution and weather variables
  summarise_at(vars(no2:boundary_layer_height), ~ mean(., na.rm = TRUE)) %>%
  # add prefix "mean_" for the relevant variables
  rename_at(vars(-c(patient_id)), ~ paste0("mean_2017_2019_", .))
```

### Computing 20 Days Exposures Lags Before the PCR

We also compute the 20 days lags of air pollution and weather exposures before the PCR test:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# create data for the 20 days lags of exposure
air_pollution_exposure_20_lags <- air_pollution_data %>%
  # select the 2020 year
  filter(year == 2020) %>%
  # merge the air pollution data with patients data
  left_join(., covid_data[, c("city", "patient_id", "date_pcr_test")], by = c("patient_id", "city")) %>%
  # compute the differences in days between exposure date and PCR date
  mutate(lags = date - date_pcr_test) %>%
  # select the 20 lags
  filter(lags >= -20 & lags <= 0) %>%
  # convert lags to positive numbers for better variable names
  mutate(lags = abs(lags)) %>%
  # compute the mean of lags for individuals with two adresses/exposures
  group_by(patient_id, city, date) %>%
  summarise_all(~ mean(., na.rm = TRUE)) %>%
  ungroup() %>%
  # pivot the data to wide format
  pivot_wider(
    id_cols = patient_id,
    names_from = lags,
    values_from = c(no2:olive),
    names_sep = "_lag_"
  )
```

# Merging the Data

We finally merge the two air pollution datasets we created with patients' Covid data:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# merge covid data with air pollution 20 days lags
final_data <-
  left_join(covid_data, air_pollution_exposure_20_lags, by = "patient_id") %>%
  # then merge with mean exposures over the 2017-2019 period
  left_join(., air_pollution_average_exposure_2017_2019, by = "patient_id") %>%
  unnest()
```

# Last Cleaning Steps

We drop Strasbourg patients with negative PCR tests for which we do not have air pollution exposure data:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
final_data <- final_data %>%
  filter(!is.na(mean_2017_2019_no2))
```

We drop patients from Bordeaux as their air pollution exposures appear to be too low to be realistic:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
final_data <- final_data %>%
  filter(city != "bordeaux")
```

We drop patients from Strasbourg who were already staying in an hospital:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
final_data <- final_data %>%
  filter(!(patient_id %in% paste("s", seq(1:106), sep = "")))
```  

We correctly label cities' names:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
final_data <- final_data %>%
  mutate(city = str_to_sentence(city),
         city = ifelse(city == "Roche", "La Roche-sur-Yon", city))
```  
  
We save the final dataset in .rds and .csv formats:

```{r, echo = TRUE, message = FALSE, warning = FALSE}
# rds format
saveRDS(final_data,
        here::here("1.data", "3.data_for_analysis", "data_analysis.rds"))
# csv format
write.csv(final_data,
          here::here("1.data", "3.data_for_analysis", "data_analysis.csv"))
```









