---
title: "Mortality & Comorbodities"
description: |
  Detailled Script.
author:
  - name: Léo Zabrocki 
    url: https://www.parisschoolofeconomics.eu/
    affiliation: Paris School of Economics
    affiliation_url: https://www.parisschoolofeconomics.eu/
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    keep_md: true
    toc: true
    toc_depth: 2
---

<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
# code chunk option
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  layout="l-body-outset",
  dev = "CairoPNG",
  dpi = 300
)
```


This script explores the relationship between COVID mortality and comorbidities. Should you have any questions or find coding errors, please do not hesitate to reach me at **leo.zabrocki@psemail.eu**.

# Required Packages and Loading Data

To reproduce exactly the `script_mortality_comorbidities.html` document, you first need to have installed:

* the [R](https://www.r-project.org/) programming language on your computer 
* [RStudio](https://rstudio.com/), an integrated development environment for R, which will allow you to knit the `script_mortality_comorbidities.Rmd` file and interact with the R code chunks
* the [R Markdown](https://rmarkdown.rstudio.com/) package
* and the [Distill](https://rstudio.github.io/distill/) package which provides the template of this document. 

Once everything is set up, we need to load the following packages:

```{r}
# load required packages
library(rmarkdown) # for creating the R Markdown document
library(knitr) # for creating the R Markdown document
library(here) # for files paths organization
library(tidyverse) # for data manipulation and visualization
library(Cairo) # for printing customed police of graphs
```

We also load our customed `ggplot2` theme for graphs:

```{r}
# load ggplot customed theme
source(
  here::here(
    "2.scripts",
    "4.custom_ggplot2_theme",
    "script_custom_ggplot_theme.R"
  )
)
```

The theme is based on the fantastic [hrbrthemes](https://hrbrmstr.github.io/hrbrthemes/index.html) package. If you do not want to use this theme or are unable to install it because of fonts issues, you can use the `theme_mimimal()` included in the `ggplot2` package.

We finally load the data we use in our analysis:

```{r}
data <-
  readRDS(here::here("1.data", "3.data_for_analysis", "data_analysis.rds"))
```

We keep patients that were COVID-positive:

```{r}
data <- data %>%
  filter(covid == 1) %>%
  dplyr::select(-covid)
```

We select patients with non-missing observations for the following variables:

```{r}
data <- data %>%
  dplyr::select(
    dead,
    sex,
    age,
    bmi,
    diabetes,
    hbp,
    active_smoking,
    copd,
    cardiac_disease_infarction,
    renal_failure,
    cancer,
    city
  ) %>%
  drop_na()
```      
      
There are `r nrow(data)` observations.

# Statistical Analysis

We run a simple logistic model to estimate the association of each patient's characteristic with the probability of dying from the COVID:

```{r}
data_regression_analysis <- data %>%
  glm(
    dead ~  sex +
      age +
      bmi +
      active_smoking +
      diabetes +
      hbp +
      copd +
      cardiac_disease_infarction +
      renal_failure +
      cancer +
      city,
    family = "binomial",
    data = .
  ) %>%
  broom::tidy(., conf.int = TRUE) %>%
  mutate_at(vars(estimate, conf.low, conf.high), ~ exp(.)) %>%
  slice(2:11) %>%
  mutate(
    term = case_when(
      term == "sexmale" ~ "Sex (Male)",
      term == "age" ~ "Age (Years)",
      term == "bmi" ~ "BMI (Kg)",
      term == "diabetes" ~ "Diabetes (Dummy)",
      term == "hbp" ~ "HBP (Dummy)",
      term == "active_smoking" ~ "Active Smoking (Dummy)",
      term == "copd" ~ "COPD (Dummy)",
      term == "cardiac_disease_infarction" ~ "Cardiac Disease Infarction (Dummy)",
      term == "renal_failure" ~ "Renal Failure (Dummy)",
      term == "cancer" ~ "Cancer (Dummy)"
    )
  )
```

Graph:

```{r, fig.width=15, fig.height=8}
# make the graph
graph_logistic_mortality_comorbidities <- data_regression_analysis %>%
  ggplot(., aes(
    x = estimate,
    y = term,
    xmin = conf.low,
    xmax = conf.high
  )) +
  geom_vline(xintercept = 1) +
  geom_pointrange(size = 1.2, colour = "deepskyblue3") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 5)) +
  ylab("") + xlab("Odds of Dying from COVID-19") +
  custom_theme +
  theme(plot.margin = margin(1, 1, 1,-1, "cm"))


# print the graph
graph_logistic_mortality_comorbidities

# save graph
ggsave(
  graph_logistic_mortality_comorbidities,
  filename = here::here(
    "3.outputs",
    "1.figures",
    "1.study_mortality_comorbidities",
    "graph_logistic_mortality_comorbidities.pdf"
  ),
  width = 30,
  height = 15,
  units = "cm",
  device = cairo_pdf
)
```

Table:

```{r}
# print the table
data_regression_analysis %>%
  select(term, estimate, conf.low, conf.high) %>%
  mutate_at(vars(estimate:conf.high), ~ round(., 1)) %>%
  rename(
    "Variable" = term,
    "Point Estimate" = estimate,
    "Lower Bound 95% CI" = conf.low,
    "Upper Bound 95% CI" = conf.high
  ) %>%
  kable(align = c("l", rep("c", 3)))
```






