---
title:  "Comorbodities & Long-Term Air Pollution Exposure"
description: |
  Detailled Script.
author:
  - name: Léo Zabrocki 
    url: https://www.parisschoolofeconomics.eu/
    affiliation: Paris School of Economics
    affiliation_url: https://www.parisschoolofeconomics.eu/
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    keep_md: true
    toc: true
    toc_depth: 2
---

<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
# code chunk option
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  layout="l-body-outset",
  dev = "CairoPNG",
  dpi = 300
)
```

# Required Packages and Loading Data

To reproduce exactly the `script_comorbidites_air_pollution.html` document, you first need to have installed:

* the [R](https://www.r-project.org/) programming language on your computer 
* [RStudio](https://rstudio.com/), an integrated development environment for R, which will allow you to knit the `script_comorbidites_air_pollution.Rmd` file and interact with the R code chunks
* the [R Markdown](https://rmarkdown.rstudio.com/) package
* and the [Distill](https://rstudio.github.io/distill/) package which provides the template of this document. 

Once everything is set up, we need to load the following packages:

```{r, echo=TRUE, message = FALSE, warning = FALSE}
# load required packages
library(rmarkdown) # for creating the R Markdown document
library(knitr) # for creating the R Markdown document
library(here) # for files paths organization
library(tidyverse) # for data manipulation and visualization
library(modelr)     # provides easy pipeline modeling functions
library(broom)      # helps to tidy up model outputs
library(lubridate) # for manipulating date variables
library(Cairo) # for printing customed police of graphs
```

We also load our custom `ggplot2` theme for graphs:

```{r}
# load ggplot custom theme
source(
  here::here(
    "2.scripts",
    "4.custom_ggplot2_theme",
    "script_custom_ggplot_theme.R"
  )
)
```

The theme is based on the fantastic [hrbrthemes](https://hrbrmstr.github.io/hrbrthemes/index.html) package. If you do not want to use this theme or are unable to install it because of fonts issues, you can use the `theme_mimimal()` included in the `ggplot2` package.

We load the data we use in our analysis:

```{r}
# load the data
data <-
  readRDS(here::here("1.data", "3.data_for_analysis", "data_analysis.rds"))
```

# Data Wrangling

We select the relevant variables:

```{r}
data <- data %>%
  dplyr::select(
    city,
    patient_id,
    dead:cancer,
    hbp,
    mean_2017_2019_no2,
    mean_2017_2019_o3,
    mean_2017_2019_pm10,
    mean_2017_2019_pm2p5,
    mean_2017_2019_so2
  )
```

We change the `sex` variable to numeric:

```{r}
data <- data %>%
  mutate(sex = ifelse(sex == "male", 1, 0))
```

Number of observations by outcome:

```{r}
# print the table
data %>%
  dplyr::select(sex, hbp, diabetes, cardiac_disease_infarction, copd, renal_failure, cancer) %>%
  summarise_all(~ sum(!is.na(.))) %>%
  pivot_longer(cols = everything(), names_to = "Comorbidity", values_to = "Number of Non-Missing Observations") %>%
  arrange(-`Number of Non-Missing Observations`) %>%
  kable(align = c("l", "c"))
```

We finally label variables:

```{r}
data <- data %>%
  # rename patients' characteristics
  rename(
    "Sex" = sex,
    "HBP" = hbp,
    "Diabetes" = diabetes,
    "Cardiac Disease Infarction" = cardiac_disease_infarction,
    "COPD" = copd,
    "Renal Failure" = renal_failure,
    "Cancer" = cancer
  ) %>%
  # rename air pollution variables
  rename(
    "NO2 Exposure (µg/m³)" = mean_2017_2019_no2,
    "O3 Exposure (µg/m³)" = mean_2017_2019_o3,
    "PM10 Exposure (µg/m³)" = mean_2017_2019_pm10,
    "PM2.5 Exposure (µg/m³)" = mean_2017_2019_pm2p5,
    "SO2 Exposure (µg/m³)" = mean_2017_2019_so2
  ) 
```

# Logistic Regression Analysis

We first run the analysis for all comorbidites but cardiac disease infarction. We run a separate analysis for this comorbidity since no patients in Reims have it.

```{r}
data_regression_analysis_wo_cdi <- data %>%
  pivot_longer(
    cols = c(`NO2 Exposure (µg/m³)`:`SO2 Exposure (µg/m³)`),
    names_to = "pollutant",
    values_to = "concentration"
  ) %>%
  pivot_longer(
    cols = c(Sex, HBP:Cancer),
    names_to = "comorbidities",
    values_to = "value_comorbidities"
  ) %>%
  filter(comorbidities != "Cardiac Disease Infarction") %>%
  group_by(pollutant, comorbidities) %>%
  nest()

data_regression_analysis_wo_cdi <-
  data_regression_analysis_wo_cdi %>%
  mutate(logistic_model = map(
    data,
    ~ glm(value_comorbidities ~ concentration + city, family = "binomial",
             data = .)
  ))

# tidy regression ouputs
data_regression_analysis_wo_cdi <-
  data_regression_analysis_wo_cdi %>%
  mutate(logistic_model = map(logistic_model, ~ broom::tidy(., conf.int = TRUE)))

# unnest results
data_regression_analysis_wo_cdi <-
  data_regression_analysis_wo_cdi %>%
  unnest(logistic_model) %>%
  dplyr::filter(term == "concentration")

# exponentiate
data_regression_analysis_wo_cdi <-
  data_regression_analysis_wo_cdi %>%
  mutate_at(vars(estimate, conf.low, conf.high), ~ exp(.))
```

We then run the analysis for cardiac disease infarction but we drop patients from Reims:

```{r}
data_regression_analysis_cdi <- data %>%
  pivot_longer(
    cols = c(`NO2 Exposure (µg/m³)`:`SO2 Exposure (µg/m³)`),
    names_to = "pollutant",
    values_to = "concentration"
  ) %>%
  pivot_longer(
    cols = c(Sex, HBP:Cancer),
    names_to = "comorbidities",
    values_to = "value_comorbidities"
  ) %>%
  filter(comorbidities == "Cardiac Disease Infarction" &
           city != "Reims") %>%
  group_by(pollutant, comorbidities) %>%
  nest()

data_regression_analysis_cdi <- data_regression_analysis_cdi %>%
  mutate(logistic_model = map(
    data,
    ~ glm(value_comorbidities ~ concentration + city, family = "binomial",
             data = .)
  ))

# tidy regression ouputs
data_regression_analysis_cdi <- data_regression_analysis_cdi %>%
  mutate(logistic_model = map(logistic_model, ~ broom::tidy(., conf.int = TRUE)))

# unnest results
data_regression_analysis_cdi <- data_regression_analysis_cdi %>%
  unnest(logistic_model) %>%
  dplyr::filter(term == "concentration")

# exponentiate
data_regression_analysis_cdi <- data_regression_analysis_cdi %>%
  mutate_at(vars(estimate, conf.low, conf.high), ~ exp(.))
```

We append the two analyses:

```{r}
data_regression_analysis <- bind_rows(data_regression_analysis_wo_cdi, data_regression_analysis_cdi)
```

Graph:

```{r, fig.width=40, fig.height=10, warning = FALSE, dev = "CairoPNG"}
# make the graph
graph_analysis_comorbidities_air_pollution <-
  data_regression_analysis %>%
  ggplot(., aes(
    x = estimate,
    y = pollutant,
    xmin = conf.low,
    xmax = conf.high
  )) +
  geom_vline(xintercept = 1) +
  geom_pointrange(size = 1.2, colour = "deepskyblue3") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 5)) +
  facet_wrap(~ comorbidities , nrow = 1) +
  ylab("") + xlab("Odds of Having the Comorbidity") +
  custom_theme +
  theme(plot.margin = margin(1, 1, 1,-1, "cm"))


# print the graph
graph_analysis_comorbidities_air_pollution

# save graph
ggsave(
  graph_analysis_comorbidities_air_pollution,
  filename = here::here(
    "3.outputs",
    "1.figures",
    "2.study_comorbidites_long_term_air_pollution",
    "graph_analysis_comorbidities_air_pollution.pdf"
  ),
  width = 50,
  height = 15,
  units = "cm",
  device = cairo_pdf
)
```


Table:

```{r}
# print the table
data_regression_analysis %>%
  dplyr::select(pollutant, comorbidities, estimate, conf.low, conf.high) %>%
  mutate_at(vars(estimate:conf.high), ~ round(., 1)) %>%
  rename(
    "Pollutant" = pollutant,
    "Comorbodities" = comorbidities,
    "Point Estimate" = estimate,
    "Lower Bound 95% CI" = conf.low,
    "Upper Bound 95% CI" = conf.high
  ) %>%
  kable(align = c("l", "l", rep("c", 3)))
```

