---
title: "Exploratory Data Analysis"
description: |
  Patients' Health Status and Air Pollution Exposures.
author:
  - name: Léo Zabrocki 
    url: https://www.parisschoolofeconomics.eu/
    affiliation: Paris School of Economics
    affiliation_url: https://www.parisschoolofeconomics.eu/
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    keep_md: true
    toc: true
    toc_depth: 2
editor_options: 
  chunk_output_type: console
---

<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
# code chunk option
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  layout="l-body-outset",
  dev = "CairoPNG",
  dpi = 300
)
```

In this script, we explore patients' health data and their air pollution exposure. Should you have any questions or find coding errors, please do not hesitate to reach me at **leo.zabrocki@psemail.eu**.

# Required Packages and Loading Data

### Required Packages

To reproduce exactly the `script_eda.html` document, you first need to have installed:

* the [R](https://www.r-project.org/) programming language on your computer 
* [RStudio](https://rstudio.com/), an integrated development environment for R, which will allow you to knit the `script_eda.Rmd` file and interact with the R code chunks
* the [R Markdown](https://rmarkdown.rstudio.com/) package
* and the [Distill](https://rstudio.github.io/distill/) package which provides the template of this document. 

Once everything is set up, we need to load the following packages:

```{r}
# load required packages
library(rmarkdown) # for creating the R Markdown document
library(knitr) # for creating the R Markdown document
library(here) # for files paths organization
library(tidyverse) # for data manipulation and visualization
library(lubridate) # for manipulating date variables
library(ggridges) # for ridge density plots
library(Cairo) # for printing custom police of graphs
```

We also load our custom `ggplot2` theme for graphs:

```{r}
# load ggplot custom theme
source(
  here::here(
    "2.scripts",
    "4.custom_ggplot2_theme",
    "script_custom_ggplot_theme.R"
  )
)
```

The theme is based on the fantastic [hrbrthemes](https://hrbrmstr.github.io/hrbrthemes/index.html) package. If you do not want to use this theme or are unable to install it because of fonts issues, you can use the `theme_mimimal()` included in the `ggplot2` package.

### Data Loading

We load the data we use in our analysis:

```{r}
data <-
  readRDS(here::here("1.data", "3.data_for_analysis", "data_analysis.rds"))
```

There are `r nrow(data)` observations and `r ncol(data)` variables. 

# Missing Data

I first display below the number of patients by city:

```{r}
# make the table
data_n_obs_city <- data %>%
  rename(City = city) %>%
  group_by(City) %>%
  summarise("Number of Patients" = n())

# print the table
data_n_obs_city %>%
  kable(align = c("l", "c"))

# save the table
write.csv(
  data_n_obs_city,
  here::here("3.outputs", "2.tables", "0.eda", "table_n_obs_city.csv")
)
```

I then display the proportion of missing data for each variable:

```{r}
# make the table
data_missing <- data %>%
  pivot_longer(
    cols = c(
      covid,
      dead,
      age,
      diabetes,
      active_smoking,
      hbp,
      cardiac_disease_infarction,
      copd,
      renal_failure,
      cancer,
      bmi
    ),
    names_to = "Variable",
    values_to = "value"
  ) %>%
  mutate(
    Variable = case_when(
      Variable == "covid" ~ "COVID",
      Variable == "dead" ~ "Dead",
      Variable == "age" ~ "Age",
      Variable == "bmi" ~ "BMI",
      Variable == "active_smoking" ~ "Active Smoking",
      Variable == "active_past_smoking" ~ "Active or Past Smoking",
      Variable == "hbp" ~ "HBP",
      Variable == "diabetes" ~ "Diabetes",
      Variable == "cardiac_disease_infarction" ~ "Cardiac Disease Infarction",
      Variable == "copd" ~ "COPD",
      Variable == "renal_failure" ~ "Renal Failure",
      Variable == "cancer" ~ "Cancer"
    )
  ) %>%
  group_by(Variable) %>%
  summarise("Missing (%)" = round(sum(is.na(value)) / n() * 100, 1))

# print the table
data_missing %>%
  arrange(-`Missing (%)`) %>%
  kable(align = c("l", "c"))

# save the table
write.csv(
  data_missing,
  here::here("3.outputs", "2.tables", "0.eda", "table_missing_data.csv")
)
```

# Health Status

### Proportion of Dead Patients

We plot the proportion of COVID positive patients who died by city:

```{r, fig.width=10, fig.height=5}
# make the graph
graph_dead_city <- data %>%
  filter(covid == 1) %>%
  group_by(city) %>%
  summarise(proportion_dead = mean(dead, na.rm = TRUE) * 100) %>%
  ggplot(., aes(x = proportion_dead, y = reorder(city, proportion_dead))) +
  geom_point(size = 6, color = "deepskyblue3") +
  xlab("Proportion of Dead Patients (%)") + ylab("") +
  custom_theme +
  theme(plot.margin = unit(c(1, 1, 1, -1), "cm"))

# print the graph
graph_dead_city

# save graph
ggsave(
  graph_dead_city,
  filename = here::here("3.outputs", "1.figures", "0.eda", "graph_dead_city.pdf"),
  width = 30,
  height = 15,
  units = "cm",
  device = cairo_pdf
)
```

### Characteristics of Patients by City

We plot the characteristics of all COVID positive patients by city:

```{r, fig.width=15, fig.height=10}
# make the graph
graph_characteristics_patients <- data %>%
  filter(covid == 1) %>%
  pivot_longer(
    cols = c(
      age,
      diabetes,
      active_smoking,
      hbp,
      cardiac_disease_infarction,
      copd,
      renal_failure,
      cancer,
      bmi
    ),
    names_to = "covariate",
    values_to = "value"
  ) %>%
  mutate(
    covariate = case_when(
      covariate == "age" ~ "Age",
      covariate == "diabetes" ~ "Diabetes (%)",
      covariate == "active_smoking" ~ "Active Smoking (%)",
      covariate == "hbp" ~ "HBP",
      covariate == "cardiac_disease_infarction" ~ "Cardiac Disease - Infarction (%)",
      covariate == "copd" ~ "COPD (%)",
      covariate == "renal_failure" ~ "Renal Failure (%)",
      covariate == "cancer" ~ "Cancer (%)",
      covariate == "bmi" ~ "BMI"
    )
  ) %>%
  group_by(city, covariate) %>%
  summarise(mean_value = mean(value, na.rm = TRUE)) %>%
  mutate(mean_value = ifelse(!(covariate %in% c("Age", "BMI")), mean_value *
                               100, mean_value)) %>%
  ggplot(., aes(x = mean_value, y = city)) +
  geom_point(size = 6, colour = "deepskyblue3") +
  facet_wrap( ~ covariate, scales = "free_x", ncol = 3) +
  xlab("") + ylab("") +
  custom_theme +
  theme(plot.margin = unit(c(1, 1, 1, -1), "cm"))

# print the graph
graph_characteristics_patients

# save graph
ggsave(
  graph_characteristics_patients,
  filename = here::here(
    "3.outputs",
    "1.figures",
    "0.eda",
    "graph_characteristics_patients.pdf"
  ),
  width = 50,
  height = 40,
  units = "cm",
  device = cairo_pdf
)
```

### Characteristics of Dead Patients

We plot the characteristics of all COVID positive patients who died by city:

```{r, fig.width=15, fig.height=10}
# make the graph
graph_characteristics_patients_mortality_outcome <- data %>%
  filter(!is.na(dead)) %>%
  filter(covid == 1) %>%
  pivot_longer(
    cols = c(
      age,
      diabetes,
      active_smoking,
      hbp,
      cardiac_disease_infarction,
      copd,
      renal_failure,
      cancer,
      bmi
    ),
    names_to = "covariate",
    values_to = "value"
  ) %>%
  mutate(
    covariate = case_when(
      covariate == "age" ~ "Age",
      covariate == "diabetes" ~ "Diabetes (%)",
      covariate == "active_smoking" ~ "Active Smoking (%)",
      covariate == "hbp" ~ "HBP",
      covariate == "cardiac_disease_infarction" ~ "Cardiac Disease - Infarction (%)",
      covariate == "copd" ~ "COPD (%)",
      covariate == "renal_failure" ~ "Renal Failure (%)",
      covariate == "cancer" ~ "Cancer (%)",
      covariate == "bmi" ~ "BMI"
    )
  ) %>%
  mutate(dead = ifelse(dead == 1, "Dead", "Alive")) %>%
  group_by(city, dead, covariate) %>%
  summarise(mean_value = mean(value, na.rm = TRUE)) %>%
  mutate(mean_value = ifelse(!(covariate %in% c("Age", "BMI")), mean_value *
                               100, mean_value)) %>%
  ggplot(., aes(x = mean_value, y = city)) +
  geom_line(aes(group = city), colour = "black") +
  geom_point(aes(colour = dead), size = 6) +
  scale_colour_manual(values = c("deepskyblue3", "tomato")) +
  facet_wrap( ~ covariate, scales = "free_x", ncol = 3) +
  xlab("") + ylab("") +
  custom_theme +
  labs(colour = "Status") +
  theme(
    legend.position = "top",
    legend.justification = "left",
    legend.direction = "horizontal",
    plot.margin = unit(c(1, 1, 1, -1), "cm")
  )

# print the graph
graph_characteristics_patients_mortality_outcome

# save graph
ggsave(
  graph_characteristics_patients_mortality_outcome,
  filename = here::here(
    "3.outputs",
    "1.figures",
    "0.eda",
    "graph_characteristics_patients_mortality_outcome.pdf"
  ),
  width = 50,
  height = 40,
  units = "cm",
  device = cairo_pdf
)
```

# Exploring Air Pollution and Weather Exposures

### Mean of Air Pollutants by City

We plot the mean concentration of each pollutant by city:

```{r fig.width=24, fig.height=8}
# make the graph
graph_mean_exposure_pollution <- data %>%
  select(city, mean_2017_2019_no2:mean_2017_2019_so2) %>%
  rename(
    "NO2" = mean_2017_2019_no2,
    "O3" = mean_2017_2019_o3,
    "PM10" = mean_2017_2019_pm10,
    "PM2.5" = mean_2017_2019_pm2p5,
    "SO2" = mean_2017_2019_so2
  ) %>%
  pivot_longer(cols = c(NO2:`SO2`),
               names_to = "pollutant",
               values_to = "concentration") %>%
  group_by(city, pollutant) %>%
  summarise(mean_concentration = mean(concentration)) %>%
  ggplot(., aes(x = mean_concentration, y = city)) +
  geom_point(size = 5, color = "deepskyblue3") +
  facet_wrap( ~ pollutant, scales = "free", nrow = 2) +
  xlab("Concentrations in (µg/m³)") + ylab("") +
  custom_theme +
  theme(plot.margin = unit(c(1, 1, 1, -1), "cm"))

# print the graph
graph_mean_exposure_pollution

# save graph
ggsave(
  graph_mean_exposure_pollution,
  filename = here::here(
    "3.outputs",
    "1.figures",
    "0.eda",
    "graph_mean_exposure_pollution.pdf"
  ),
  width = 40,
  height = 30,
  units = "cm",
  device = cairo_pdf
)
```

### Air Pollutants Density Distributions

We plot the concentration distribution of each pollutant by city:

```{r fig.width=24, fig.height=8}
# make the graph
graph_density_pollution_exposure <- data %>%
  select(city, mean_2017_2019_no2:mean_2017_2019_so2) %>%
  rename(
    "NO2" = mean_2017_2019_no2,
    "O3" = mean_2017_2019_o3,
    "PM10" = mean_2017_2019_pm10,
    "PM2.5" = mean_2017_2019_pm2p5,
    "SO2" = mean_2017_2019_so2
  ) %>%
  pivot_longer(cols = c(NO2:`SO2`),
               names_to = "pollutant",
               values_to = "concentration") %>%
  ggplot(., aes(x = concentration, y = city)) +
  geom_density_ridges(colour = NA,
                      fill = "deepskyblue3",
                      alpha = 0.8) +
  facet_wrap( ~ pollutant, scales = "free", nrow = 1) +
  xlab("Concentration (µg/m3) ") + ylab("") +
  custom_theme +
  theme(plot.margin = unit(c(1, 1, 1, -1), "cm"))

# print the graph
graph_density_pollution_exposure

# save graph
ggsave(
  graph_density_pollution_exposure,
  filename = here::here(
    "3.outputs",
    "1.figures",
    "0.eda",
    "graph_density_pollution_exposure.pdf"
  ),
  width = 70,
  height = 15,
  units = "cm",
  device = cairo_pdf
)
```

### Weather Parameters Density Distributions

We plot the density distribution of each weather covariate by city:

```{r fig.width=27, fig.height=8}
# make the graph
graph_density_exposure_weather <- data %>%
  select(city,
         mean_2017_2019_average_temperature:mean_2017_2019_boundary_layer_height) %>%
  rename(
    "Average Temperature (°C)" = mean_2017_2019_average_temperature,
    "Wind Speed (m/s)" = mean_2017_2019_wind_speed,
    "Boundary Layer Height (m)" = mean_2017_2019_boundary_layer_height,
    "Relative Humidity (%)" = mean_2017_2019_relative_humidity
  ) %>%
  pivot_longer(
    cols = c(`Average Temperature (°C)`:`Boundary Layer Height (m)`),
    names_to = "weather_parameter",
    values_to = "value"
  ) %>%
  ggplot(., aes(x = value, y = city)) +
  geom_density_ridges(colour = NA,
                      fill = "deepskyblue3",
                      alpha = 0.8) +
  facet_wrap(~ weather_parameter, scales = "free", nrow = 1) +
  xlab("Value") + ylab("") +
  custom_theme +
  theme(plot.margin = unit(c(1, 1, 1, -1), "cm"))

# print the graph
graph_density_exposure_weather

# save graph
ggsave(
  graph_density_exposure_weather,
  filename = here::here(
    "3.outputs",
    "1.figures",
    "0.eda",
    "graph_density_exposure_weather.pdf"
  ),
  width = 62,
  height = 18,
  units = "cm",
  device = cairo_pdf
)
```


# Air Pollution - Death Relationship

We plot below the average concentration of each pollutant by mortality outcome and city:

```{r fig.width=15, fig.height=7}
# make the graph
graph_mortality_mean_exposure <- data %>%
  filter(!is.na(dead)) %>%
  filter(covid == 1) %>%
  rename(
    "NO2" = mean_2017_2019_no2,
    "O3" = mean_2017_2019_o3,
    "PM10" = mean_2017_2019_pm10,
    "PM2.5" = mean_2017_2019_pm2p5,
    "SO2" = mean_2017_2019_so2
  ) %>%
  pivot_longer(cols = c(NO2:`SO2`),
               names_to = "pollutant",
               values_to = "concentration") %>%
  mutate(dead = ifelse(dead == 1, "Dead", "Alive")) %>%
  group_by(city, dead, pollutant) %>%
  summarise(mean_concentration = mean(concentration, na.rm = TRUE)) %>%
  ggplot(., aes(x = mean_concentration, y = city)) +
  geom_line(aes(group = city), colour = "black") +
  geom_point(aes(colour = dead), size = 4) +
  scale_colour_manual(values = c("deepskyblue3", "tomato")) +
  facet_wrap( ~ pollutant, scales = "free_x", ncol = 5) +
  xlab("Concentration (µg/m3)") + ylab("") +
  custom_theme +
  labs(colour = "Status") +
  theme(
    legend.position = "top",
    legend.justification = "left",
    legend.direction = "horizontal",
    plot.margin = unit(c(1, 1, 1, -1), "cm")
  )

# print the graph
graph_mortality_mean_exposure

# save graph
ggsave(
  graph_mortality_mean_exposure,
  filename = here::here(
    "3.outputs",
    "1.figures",
    "0.eda",
    "graph_mortality_mean_exposure.pdf"
  ),
  width = 60,
  height = 15,
  units = "cm",
  device = cairo_pdf
)
```

# Air Pollution - Comorbidities

We plot below the average concentration of each pollutant by co-morbity:

```{r fig.width=20, fig.height=7}
# make the graph
graph_pollution_exposure_comorbidites <- data %>%
  dplyr::select(
    hbp:diabetes,
    cardiac_disease_infarction:cancer,
    mean_2017_2019_no2:mean_2017_2019_so2
  ) %>%
  drop_na() %>%
  mutate_at(vars(hbp:cancer), ~ as.character(.)) %>%
  pivot_longer(cols = c(hbp:cancer),
               names_to = "comorbidity",
               values_to = "value") %>%
  mutate(
    comorbidity = case_when(
      comorbidity == "hbp" ~ "HBP",
      comorbidity == "diabetes" ~ "Diabetes (%)",
      comorbidity == "cardiac_disease_infarction" ~ "Cardiac Disease - Infarction (%)",
      comorbidity == "copd" ~ "COPD (%)",
      comorbidity == "renal_failure" ~ "Renal Failure (%)",
      comorbidity == "cancer" ~ "Cancer (%)",
      comorbidity == "bmi" ~ "BMI"
    )
  ) %>%
  group_by(comorbidity, value) %>%
  summarise_at(vars(mean_2017_2019_no2:mean_2017_2019_so2), ~ mean(.)) %>%
  rename(
    "NO2" = mean_2017_2019_no2,
    "O3" = mean_2017_2019_o3,
    "PM10" = mean_2017_2019_pm10,
    "PM2.5" = mean_2017_2019_pm2p5,
    "SO2" = mean_2017_2019_so2
  ) %>%
  pivot_longer(cols = c(NO2:`SO2`),
               names_to = "pollutant",
               values_to = "concentration") %>%
  mutate(value = ifelse(value == 1, "Has the disease", "Does not have the disease")) %>%
  ggplot(., aes(x = concentration, y = comorbidity)) +
  geom_line(aes(group = comorbidity), colour = "black") +
  geom_point(aes(colour = value), size = 4) +
  scale_colour_manual(values = c("deepskyblue3", "tomato")) +
  facet_wrap( ~ pollutant, scales = "free_x", ncol = 5) +
  xlab("Concentration (µg/m3) ") + ylab("") +
  custom_theme +
  labs(colour = "Status") +
  theme(
    legend.position = "top",
    legend.justification = "left",
    legend.direction = "horizontal",
    plot.margin = unit(c(1, 1, 1, -1), "cm")
  )

# print the graph
graph_pollution_exposure_comorbidites

# save graph
ggsave(
  graph_pollution_exposure_comorbidites,
  filename = here::here(
    "3.outputs",
    "1.figures",
    "0.eda",
    "graph_pollution_exposure_comorbidites.pdf"
  ),
  width = 60,
  height = 15,
  units = "cm",
  device = cairo_pdf
)
```

We plot below the average concentration of each pollutant by co-morbity and city:

```{r fig.width=30, fig.height=30}
# make the graph
graph_pollution_exposure_comorbidites_city <- data %>%
  dplyr::select(
    city,
    hbp:diabetes,
    cardiac_disease_infarction:cancer,
    mean_2017_2019_no2:mean_2017_2019_so2
  ) %>%
  mutate(city = ifelse(city == "La Roche-sur-Yon", "La Roche-\n sur-Yon", city)) %>%
  drop_na() %>%
  mutate_at(vars(hbp:cancer), ~ as.character(.)) %>%
  pivot_longer(cols = c(hbp:cancer),
               names_to = "comorbidity",
               values_to = "value") %>%
  mutate(
    comorbidity = case_when(
      comorbidity == "hbp" ~ "HBP",
      comorbidity == "diabetes" ~ "Diabetes (%)",
      comorbidity == "cardiac_disease_infarction" ~ "Cardiac Disease - Infarction (%)",
      comorbidity == "copd" ~ "COPD (%)",
      comorbidity == "renal_failure" ~ "Renal Failure (%)",
      comorbidity == "cancer" ~ "Cancer (%)",
      comorbidity == "bmi" ~ "BMI"
    )
  ) %>%
  group_by(city, comorbidity, value) %>%
  summarise_at(vars(mean_2017_2019_no2:mean_2017_2019_so2), ~ mean(.)) %>%
  rename(
    "NO2" = mean_2017_2019_no2,
    "O3" = mean_2017_2019_o3,
    "PM10" = mean_2017_2019_pm10,
    "PM2.5" = mean_2017_2019_pm2p5,
    "SO2" = mean_2017_2019_so2
  ) %>%
  pivot_longer(cols = c(NO2:`SO2`),
               names_to = "pollutant",
               values_to = "concentration") %>%
  mutate(value = ifelse(value == 1, "Has the disease", "Does not have the disease")) %>%
  ggplot(., aes(x = concentration, y = comorbidity)) +
  geom_line(aes(group = comorbidity), colour = "black") +
  geom_point(aes(colour = value), size = 4) +
  scale_colour_manual(values = c("deepskyblue3", "tomato")) +
  facet_grid(city ~ pollutant, scales = "free") +
  xlab("Concentration (µg/m3) ") + ylab("") +
  custom_theme +
  labs(colour = "Status") +
  theme(
    legend.position = "top",
    legend.justification = "left",
    legend.direction = "horizontal",
    plot.margin = unit(c(1, 1, 1, -1), "cm")
  )

# print the graph
graph_pollution_exposure_comorbidites_city

# save graph
ggsave(
  graph_pollution_exposure_comorbidites_city,
  filename = here::here(
    "3.outputs",
    "1.figures",
    "0.eda",
    "graph_pollution_exposure_comorbidites_city.pdf"
  ),
  width = 50,
  height = 50,
  units = "cm",
  device = cairo_pdf
)
```