---
title: "COVID-19 Mortality & Short-Term Exposure to Environmental Factors"
description: |
  Detailled Script.
author:
  - name: Léo Zabrocki 
    url: https://www.parisschoolofeconomics.eu/
    affiliation: Paris School of Economics
    affiliation_url: https://www.parisschoolofeconomics.eu/
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    keep_md: true
    toc: true
    toc_depth: 2
---

<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
# code chunk option
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  layout="l-body-outset",
  dev = "CairoPNG",
  dpi = 300
)
```

In this document, we explore the relationship between short-term environmental factors exposure (averaged the 20 days before the PCR test) and COVID-19 mortality. Should you have any questions or find coding errors, please do not hesitate to reach me at **leo.zabrocki@psemail.eu**.

# Required Packages and Loading Data

To reproduce exactly the `script_mortality_short_term_environmental_exposure.html` document, you first need to have installed:

* the [R](https://www.r-project.org/) programming language on your computer 
* [RStudio](https://rstudio.com/), an integrated development environment for R, which will allow you to knit the `script_mortality_short_term_environmental_exposure.Rmd` file and interact with the R code chunks
* the [R Markdown](https://rmarkdown.rstudio.com/) package
* and the [Distill](https://rstudio.github.io/distill/) package which provides the template of this document. 

Once everything is set up, we need to load the following packages:

```{r}
# load required packages
library(rmarkdown) # for creating the R Markdown document
library(knitr) # for creating the R Markdown document
library(here) # for files paths organization
library(tidyverse) # for data manipulation and visualization
library(modelr) # provides easy pipeline modeling functions
library(broom) # helps to tidy up model outputs
library(lubridate) # for manipulating date variables
library(Cairo) # for printing customed police of graphs
```

We also load our custom `ggplot2` theme for graphs:

```{r}
# load ggplot custom theme
source(
  here::here(
    "2.scripts",
    "4.custom_ggplot2_theme",
    "script_custom_ggplot_theme.R"
  )
)
```

The theme is based on the fantastic [hrbrthemes](https://hrbrmstr.github.io/hrbrthemes/index.html) package. If you do not want to use this theme or are unable to install it because of fonts issues, you can use the `theme_mimimal()` included in the `ggplot2` package.

We load the data we use in our analysis:

```{r}
data <-
  readRDS(here::here("1.data", "3.data_for_analysis", "data_analysis.rds"))
```

# Data Wrangling

### Data Cleaning

We select the relevant variables:

```{r}
data <- data %>%
  mutate(month = lubridate::month(date_pcr_test) %>% as.factor(.)) %>%
  dplyr::select(
    city,
    patient_id,
    month,
    covid:cancer,
    no2_lag_20:wind_speed_lag_0,
    uvi_lag_20:olive_lag_0
  )
```

We keep patients that were COVID-positive:

```{r}
data <- data %>%
  filter(covid == 1) %>%
  dplyr::select(-covid)
```

We compute the average exposure for an environmental factor over 20 days before the PCR test:

```{r}
# compute average of short-term exposure
data_short_term_exposure <- data %>%
  select(patient_id, no2_lag_20:olive_lag_0) %>%
  pivot_longer(
    cols = c(no2_lag_20:olive_lag_0),
    names_to = "exposure_lag",
    values_to = "value"
  ) %>%
  mutate(
    exposure = case_when(
      str_detect(exposure_lag, "no2") ~ "NO2 (µg/m³)",
      str_detect(exposure_lag, "o3") ~ "O3 (µg/m³)",
      str_detect(exposure_lag, "pm10") ~ "PM10 (µg/m³)",
      str_detect(exposure_lag, "pm2p5") ~ "PM2.5 (µg/m³)",
      str_detect(exposure_lag, "so2") ~ "SO2 (µg/m³)",
      str_detect(exposure_lag, "temperature") ~ "Average Temperature (°C)",
      str_detect(exposure_lag, "humidity") ~ "Average Humidity (%)",
      str_detect(exposure_lag, "wind") ~ "Wind Speed (m/s)",
      str_detect(exposure_lag, "uvi") ~ "UV",
      str_detect(exposure_lag, "birch") ~ "Birch Pollen (grains/m³)",
      str_detect(exposure_lag, "grass") ~ "Grass Pollen (grains/m³)",
      str_detect(exposure_lag, "olive") ~ "Olive Pollen (grains/m³)"
    )
  ) %>%
  group_by(patient_id, exposure) %>%
  summarise(mean_exposure = mean(value)) %>%
  pivot_wider(names_from = exposure, values_from = mean_exposure)
```

We merge back these exposures with the inital data:

```{r}
data <- data %>%
  select(city:cancer) %>%
  left_join(., data_short_term_exposure, by = "patient_id")
```

# Logistic Regression Analysis

We run our multivariate logistic regression model:

```{r}
data_regression_analysis <- data %>%
  pivot_longer(
    cols = c(`Average Humidity (%)`:`Wind Speed (m/s)`),
    names_to = "exposure",
    values_to = "exposure_value"
  ) %>%
  group_by(exposure) %>%
  nest()

data_regression_analysis <- data_regression_analysis %>%
  mutate(
    # model
    model= map(
      data,
      ~ glm(
        dead ~ exposure_value +
          sex +
          age +
          bmi +
          diabetes +
          hbp +
          copd +
          cardiac_disease_infarction +
          renal_failure +
          cancer +
          month +
          city,
        data = .
      )
    )
  )

# tidy regression ouputs
data_regression_analysis <- data_regression_analysis %>%
  select(-data) %>%
  mutate(models_dfs = map(model, ~ broom::tidy(., conf.int = TRUE)))

# unnest results and select coefficient
data_regression_analysis <- data_regression_analysis %>%
  unnest(models_dfs) %>%
  dplyr::filter(term == "exposure_value")

# exponentiate
data_regression_analysis <- data_regression_analysis %>%
  mutate_at(vars(estimate, conf.low, conf.high), ~ exp(.))
```

We graphically display point estimates and their associated 95% confidence intervals:

```{r, fig.width=20, fig.height=10}
# make the graph
graph_logistic_all_cities <- data_regression_analysis %>%
  filter(exposure != "Olive Pollen (grains/m³)") %>%
  ggplot(., aes(
    x = estimate,
    y = fct_rev(exposure),
    xmin = conf.low,
    xmax = conf.high
  )) +
  geom_vline(xintercept = 1) +
  geom_pointrange(size = 1.2, colour = "deepskyblue3") +
  ylab("") + xlab("Odds of Dying from COVID-19") +
  custom_theme +
  theme(plot.margin = margin(1, 1, 1, -1, "cm"))


# print the graph
graph_logistic_all_cities

# save graph
ggsave(graph_logistic_all_cities, filename = here::here("3.outputs", "1.figures", "4.study_mortality_short_term_environmental_exposure", "graph_mortality_short_term_environmental_exposure.pdf"), 
       width = 30, height = 20, units = "cm", device = cairo_pdf)
```

We display values of point estimates and their associated 95% confidence intervals in the table below:

```{r}
# print the table
options(scipen = 100)

data_regression_analysis %>%
  filter(exposure != "Olive Pollen (grains/m³)") %>%
  select(exposure, estimate, conf.low, conf.high) %>%
  mutate_at(vars(estimate:conf.high), ~ round(., 2)) %>%
  rename("Exposure" = exposure, "Point Estimate" = estimate, "Lower Bound 95% CI" = conf.low, "Upper Bound 95% CI" = conf.high) %>%
  kable(align = c("l", rep("c", 3)))
```